---
title: "Project Proposal"
author: "Team name - Team member 1, Team member 2, Team member 3, Team member 4"
format: pdf
execute: 
  warning: false
  message: false
editor: visual
---

```{r}
#| label: load-pkg-data
library(tidyverse)
library(tidymodels)
# add other packages as needed

```

## Introduction

...

## Data description

...

## Exploratory data analysis

```{r}
#| label: EDA-mutate-certificate

imdb_top_1000 <- imdb_top_1000 %>% 
  mutate(no_votes_scaled = No_of_Votes / 10^6,
         gross_scaled = Gross / 10^6,
         Certificate = case_when(Certificate == "U" ~ "U", 
                          Certificate == "A" ~ "A",
                          TRUE ~ "Other"))

```

```{r}
#| label: visualizations

imdb_top_1000 %>% 
  ggplot(aes(x = no_votes_scaled, y = Meta_score)) +
    geom_point() +
    labs(x = "Number of Votes (In Millions)",
         y = "Meta-Score",
         title = "Relationship Between Number of Votes on IMDB and Critic Meta-Score")

imdb_top_1000 %>% 
  ggplot(aes(x = gross_scaled, y = Meta_score)) +
    geom_point() +
    labs(x = "Gross Revenue (In Millions)",
         y = "Meta-Score",
         title = "Relationship Between Gross Revenue and Critic Meta-Score")

imdb_top_1000 %>%
  ggplot(aes(x = Certificate)) +
  geom_bar() +
  labs(x = "Certificate",
           y = "Frequency",
           title = "Distribution of Types of Certificate in Films")

imdb_top_1000 %>% 
  ggplot(aes(x = no_votes_scaled, y = IMDB_Rating)) +
    geom_point()+
    labs(x = "Number of IMDB Votes (In Millions)",
         y = "IMDB Rating",
         title = "Relationship Between Number of Votes on IMDB Ratings")

imdb_top_1000 %>% 
  ggplot(aes(x = gross_scaled, y = IMDB_Rating)) +
    geom_point() +
    labs(x = "Gross Revenue (In Millions)",
         y = "IMDB Rating",
         title = "Relationship Between Gross Revenue and IMDB Ratings")
```

Upon initial EDA, it seems as if the relationship between gross revenue as a predictor for both IMDB and Metascore ratings are both nonlinear. Likewise, we see the same for the number of votes in relationship with the response variables. It does seem that there is a positive relationship between the number of votes on IMDB and the IMDB rating, but that is to be expected. For the other relationships, we may need to consider doing some data transformation.

```{r}
#| label: interaction effects

imdb_top_1000 %>% 
  ggplot(aes(x = gross_scaled, y = Meta_score, color = Certificate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

```

Upon initial EDA, it seems as if there is an interaction effect based on certificate of the film between films' gross revenue and their Meta score.


```{r}
#| label: response distribution
imdb_top_1000 %>% 
ggplot(aes(x = IMDB_Rating)) +
  geom_histogram()
```
The distribution of IMDB Ratings (one of our response variables) seems to be skewed right.

```{r}
#| label: response-dist
imdb_top_1000 %>% 
ggplot(aes(x = Meta_score)) +
  geom_histogram()
```
The distribution of meta-score (our other response variable) seems to be skewed left.

```{r}
#| label: meta-summary-stats

imdb_top_1000 %>%
  summarise(
    meta_mean = mean(Meta_score),
    meta_median = median(Meta_score),
    meta_sd = sd(Meta_score),
    meta_IQR = IQR(Meta_score),
    meta_min = min(Meta_score),
    meta_max = max(Meta_score)
  )
```

## Analysis approach

...

## Data dictionary

The data dictionary can be found [here](ADD%20LINK!!!) \[Update the link and remove this note!\]
